---
title: "Reproducible Research Assignment 1"
author: "Mahbub khan"
date: "April 6, 2016"
output: html_document
---



##Objective:
This assignment analyses "activity monitoring data", looking at total numbers of activity each day and average daily activity pattern. It also looks at the effect of missing values and effect of imputing them by the mean activity rate.Lastly, it looks for pattern in activity divided into weekdays and weekends. 


##Loading and Preprocessing data
```{r, echo=TRUE}
actvdata<-read.csv("./data/activity.csv",header = TRUE, sep = ",")#loading csv file from data folder of home drive
str(actvdata)
actvdata$date<-as.Date(actvdata$date)#formatting as date
str(actvdata$date)
```

##Total Number of Steps per day


```{r,}
library(dplyr)#loading dplyr package
bydate<-summarise(group_by(actvdata,date),totalsteps=sum(steps))#getting total steps by date
head(bydate)


hist(bydate$totalsteps)#making histogram of total steps by date
mean(bydate$totalsteps,na.rm = TRUE)#mean of total stpes
median(bydate$totalsteps,na.rm = TRUE)#median of total steps
```

##Average Daily Activity Pattern


```{r,}
intstepsdf<-actvdata[,c(3,1) ]#subsetting a datframe with interval and steps only
head(intstepsdf)
byinterval<-aggregate(.~interval, data=intstepsdf, mean)# mean steps by interval
head(byinterval)
plot(byinterval$steps~byinterval$interval,type="l")#plotting timeseries of mean steps
byinterval[byinterval$steps==max(byinterval$steps),]#max mean step by interval

```

##Imputing Missing Values


```{r,}
sum(is.na(actvdata$steps))#total NA's


for (i in 1:nrow(actvdata)){
  if(is.na(actvdata$steps[i])){
    actvdata$steps[i]<-round(byinterval[byinterval$interval==actvdata$interval[i],2])}}#replacing NA's of steps in actvdata with mean steps of same interval from byinterval df

sum(is.na(actvdata$steps))#checking for NA's


clnbydate<-summarise(group_by(actvdata,date),totalsteps=sum(steps))#getting total steps by date of NA imputed data

hist(clnbydate$totalsteps)#histogram of cleaned steps
mean(clnbydate$totalsteps)# mean of cleaned steps
median(clnbydate$totalsteps)#median of cleaned steps

```

###Impact of Imputing Missing Data

Looking at the two histograms we can see that, the middle bin of 10000 to 15000 is taller for the cleaned data, and this is due to inputting more values of that range instead of missing values. We also see that median and mean becomes same for cleaned data, this due to having a lot of values with mean value, hence median the middle value becomes one of them.




##Difference between Weekend and Weekday Activity Pattern
```{r,}
weekdays(clnbydate$date[1:10])#creating a new column "days"(Weekday & Weekend) in actvdata df
for (i in 1:nrow(actvdata)){
  if(weekdays(actvdata$date[i])=="Saturday"|weekdays(actvdata$date[i])=="Sunday"){
    actvdata$days[i]<-"Weekend"}
  else {actvdata$days[i]<-"Weekday"
    
  }
}
head(actvdata)


intdayssteps<-summarise(group_by(actvdata,interval,days),meansteps=round(mean(steps)))#geting the mean steps for cleaned data
head(intdayssteps)
library(ggplot2)#loading ggplot2 package

p<- ggplot(intdayssteps, aes(x=interval, y=meansteps, color = days)) +
  geom_line() +
  facet_wrap(~days, ncol = 1, nrow=2)#creating the weekend/weekday activity timeseries
print(p)#plotting it


```

###Interpretation of the plot
From the time series plots of activity for weekdays and weekends, we can see that the subject is more active during weekends, but the highest activity happens on a weekday.
